---
title: "Programmatic Sampling Pipeline"
subtitle: >
  Reproducible spatial sampling, GPS point management, and static mapping
  without the Shiny app.
output:
  rmarkdown::html_vignette:
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Programmatic Sampling Pipeline}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The `gpssampling` package includes a **programmatic sampling pipeline** that
runs entirely in R scripts -- no Shiny app required. This is useful for:

- Reproducible, scripted sampling workflows
- Batch processing multiple survey areas
- Integration with existing analysis pipelines
- Generating publication-quality static maps

The pipeline has three stages:

1. **Sampling** -- fetch buildings, filter, crop to communities, and sample
   with minimum-distance constraints.
2. **GPS Management** -- split into batches, create buffers, export to GPX /
   GeoPackage / SQLite tile overlays, zip, and email.
3. **Static Mapping** -- generate per-community and overview maps with
   basemap tiles, batch coloring, and scale bars.

## Stage 1: Sampling

### Fetch and filter buildings

```{r fetch}
library(gpssampling)
library(sf)

# Load your study area boundary and community polygons
state_boundary <- st_read("boundary.gpkg")
communities <- st_read("communities.gpkg")

# Download OSM building footprints
osm_buildings <- fetch_osm_buildings(state_boundary)

# Filter out non-residential buildings (hospitals, schools, etc.)
buildings <- filter_buildings(osm_buildings)
```

If you have your own building footprint dataset (e.g., from Microsoft Building
Footprints or Google Open Buildings), use OSM only for labeling:

```{r user-buildings}
user_buildings <- st_read("my_building_footprints.gpkg")
osm <- fetch_osm_buildings(state_boundary)

buildings <- filter_buildings(
  user_buildings,
  osm_buildings_sf = osm,
  remove_tags = c("hospital", "school", "church")
)
```

### Crop to communities

`crop_buildings()` intersects buildings with community polygons, converts to
centroids, and returns a named list (one entry per community):

```{r crop}
buildings_list <- crop_buildings(
  buildings,
  communities,
  community_id_col = "name"
)

# Check counts per community
vapply(buildings_list, nrow, integer(1L))
#> community_one community_two community_three community_four
#>           847          1203             956            612
```

### Visualize cropped buildings

`map_cropped_buildings()` produces one large map per community showing the
community boundary and actual building footprint shapes over an OSM basemap.
Pass the original building polygons -- the function clips them to each
community internally:

```{r map-crop}
crop_maps <- map_cropped_buildings(
  buildings,
  communities,
  community_id_col = "name",
  out_dir = "output/maps"
)
# Or save individually:
# ggplot2::ggsave("community_one.png", crop_maps[["community_one"]],
#   width = 12, height = 12, dpi = 300)
```

### Sample with minimum-distance constraints

`sample_communities()` selects points using a two-phase approach:

1. **Random selection** -- all points (including the first) are chosen at
   random, subject to the `min_distance` constraint so that no two selected
   points are closer than `min_distance` meters.
2. **Proximity ordering** -- after selection, the points are reordered for
   efficient field work: the selected point closest to an OSM road becomes
   #1, then the nearest unvisited selected point becomes #2, and so on
   (greedy nearest-neighbour chain). This `selection_order` determines batch
   assignment and the sequence field workers follow on the ground.

```{r sample}
samples <- sample_communities(
  buildings_list,
  n_required = c(
    community_one = 30, community_two = 80,
    community_three = 85, community_four = 60
  ),
  min_distance = 50,
  seed = 250292L
)
```

The result is a named list of lists. Each community contains `$primary`
(selected points with `selection_order`), `$secondary` (remaining points),
`$buildings` (all candidates), and metadata (`$min_distance`, `$seed`).

**Reproducibility:** Same `seed` + same inputs = same output on any platform.
The algorithm uses `withr::with_seed()` (not `set.seed()`), deterministic
point sorting, and alphabetical community processing order.

## Stage 2: GPS Point Management

### Split into batches

Distribute points across field teams using round-robin assignment:

```{r batches}
batched <- split_batches(samples, n_batches = 5L, set = "primary")
```

Each community's points get an `assigned_batch` column (1 to 5).

### Create buffers

Generate circular buffer polygons around sampled points:

```{r buffers}
buffers <- create_buffers(samples, radius = 50, set = "primary")
```

### Export to disk

`export_points()` writes points, buffer polygons, and OsmAnd-compatible
SQLite tile overlays to a structured directory:

```{r export}
manifest <- export_points(
  batched,
  out_dir = "output",
  formats = c("gpkg", "gpx"),
  include_buffers = TRUE,
  buffer_radius = 50,
  set = "primary"
)
```

This creates:

```
output/
  primary/
    community_one/
      community_one_primary_all.gpkg
      community_one_primary_batch_1.gpkg
      community_one_primary_batch_1.gpx
      community_one_buffers_all.gpkg
      community_one_buffers_all.gpx
      community_one_buffers_all.sqlitedb      <- OsmAnd tile overlay
      community_one_buffers_batch_1.gpkg
      community_one_buffers_batch_1.gpx
      community_one_buffers_batch_1.sqlitedb
      ...
    community_two/
      ...
```

**OsmAnd integration:** Copy `.sqlitedb` files to
`Android/data/net.osmand/files/tiles/` on the tablet and enable via
Configure map > Overlay/Underlay.

### Zip and email

```{r zip-email}
# Zip GPX + SQLite files for field distribution
zips <- zip_points("output", prefix = "project-", sets = "primary")

# Send via Outlook (requires emayili package + SMTP credentials)
email_points(
  zips,
  to = "fieldteam@example.org",
  subject = "Survey - GPS Sampling Points"
)
```

## Stage 3: Static Mapping

Generate publication-quality static maps using `ggplot2` + basemap tiles:

```{r maps}
# Generate all maps (overview + per-community for primary and secondary)
maps <- map_all_communities(
  samples,
  communities,
  community_id_col = "name",
  out_dir = "output/maps",
  buffer_radius = 50
)
```

This creates:

- `overview.png` -- all communities, uniform point color, community labels
- `{name}_primary.png` -- batch-colored primary points per community
- `{name}_secondary.png` -- batch-colored secondary points per community

You can also generate individual maps:

```{r single-map}
# Single community map
p <- map_community(
  "community_one",
  communities[communities$name == "community_one", ],
  samples$community_one$primary,
  buffers_sf = create_buffers(samples$community_one$primary, radius = 50),
  batch_colors = TRUE
)

# Customize before saving
p + ggplot2::theme(legend.position = "bottom")
ggplot2::ggsave("community_one.png", p, width = 10, height = 12, dpi = 300)
```

## Complete Workflow

```{r full-pipeline}
library(gpssampling)
library(sf)

# 1. Load data
communities <- st_read("communities.gpkg")
state_boundary <- st_read("boundary.gpkg")

# 2. Fetch, filter, crop buildings
buildings_list <- fetch_osm_buildings(state_boundary) |>
  filter_buildings() |>
  crop_buildings(communities, community_id_col = "name")

# 3. Sample (seed is required — choose any integer for reproducibility)
samples <- sample_communities(
  buildings_list,
  n_required = c(
    community_one = 30, community_two = 80,
    community_three = 85, community_four = 60
  ),
  min_distance = 50,
  seed = 250292L
)

# 4. Batch and export
batched <- split_batches(samples, n_batches = 5L, set = "primary")
export_points(batched, out_dir = "output", set = "primary")
export_points(
  split_batches(samples, n_batches = 5L, set = "secondary"),
  out_dir = "output", set = "secondary"
)

# 5. Zip for field teams
zips <- zip_points("output", prefix = "project-")

# 6. Generate maps
map_all_communities(
  samples, communities,
  community_id_col = "name",
  out_dir = "output/maps"
)
```

## Function Reference

### Sampling

| Function | Purpose |
|----------|---------|
| `fetch_osm_buildings()` | Download OSM building footprints for an area |
| `filter_buildings()` | Remove non-residential buildings by OSM tags |
| `crop_buildings()` | Clip buildings to community polygons, convert to centroids |
| `find_start_points()` | Find closest-to-road building per community (utility) |
| `sample_communities()` | Random sampling with proximity-based ordering |

### GPS Management

| Function | Purpose |
|----------|---------|
| `split_batches()` | Assign round-robin batch numbers to points |
| `create_buffers()` | Generate circular buffer polygons |
| `create_buffer_tiles()` | Create OsmAnd-compatible SQLite tile overlays |
| `export_points()` | Export points, buffers, and tiles to disk |
| `zip_points()` | Bundle GPX + SQLite files into zip archives |
| `email_points()` | Send zip files via SMTP email |

### Static Mapping

| Function | Purpose |
|----------|---------|
| `map_cropped_buildings()` | Side-by-side panels of cropped buildings per community |
| `map_community()` | Per-community map with batch coloring |
| `map_overview()` | Zoomed-out overview of all communities |
| `map_all_communities()` | Generate and optionally save all maps |
